How economic, health and social factors affect life expectancy in developing countries?

### 2025-04-10 paper.Rmd
added remendations section in conclusion

```
The first transformation we had was with health expenditure as a percentage of GDP per capita. For health expenditure, the figure above reveals a non-linear, positive exponential relationship with life expectancy. To linearize this relationship, we applied a $log(1+Health Expenditure)$ transformation as the $log$ function is in the inverse of exponential. We used $log(1+x)$ rather than $log(x)$ to accommodate the theoretical possibility of zero expenditure values since $log(0$) is undefined.

@@

```
For infant deaths, we applied a $log(1+\text{infant deaths})$ transformation based on the weak exponential relationship observed in the figure. We used $log(1+x)$ to maintain a valid domain for potential zero values. The transformed data reveals a generally negative relationship with life expectancy, though with two distinct patterns: higher variability and occasional uniformity between 0-2, and a curved downward slope between 2-6. After testing multiple transformations, $log(1+x)$ yielded the best results, despite not resulting in a perfectly linear relationship.

We also considered removing outlier countries with over 900 infant deaths per 100 people as shown in the figure. This was done to potentially create a stronger exponential relationship between infant deaths and life expectancy. However, after removing the outliers, the transformation still showed no improved linearity after removal. Furthermore, the final model results remained largely unchanged without these outliers, indicating they were not influential. Therefore, we kept these data points in the final model dataset.

@@

```
For the third and final transformation, we decided to do a $log(1+\text{HIV/AIDS Deaths})$ transformation. From the figure above, the relationship between aids and life expectancy is negative and exponential so a log transformation is valid. Similarly, $log(1 + x)$ is used to ensure a valid domain. From the transformation plot above, $log(1+\text{HIV/AIDS Deaths})$ and life expectancy appear to have a moderately strong negative linear relationship.

@@

In this study we explored factors that influence life expectancy in developing countries using multiple linear regression. We obtained our dataset from Kaggle (Life Expectancy WHO dataset), which compiles health, demographic, and economic indicators from the World Health Organization and United Nations. After cleaning the dataset and removing missing values, we focused on five predictors: schooling, alcohol consumption (categorical), infant deaths, HIV/AIDS prevalence, and health expenditure (as a percentage of GDP per capita). Because the relationships for infant deaths, HIV/AIDS, and health expenditure with life expectancy were all exponential-like, log-transformations were applied to stabilize variance and improve model fit. Residual analysis confirmed no major violations of linear regression model assumptions, while model selection using regsubsets confirmed that our final set of predictors balanced model simplicity and predictive strength. Our goal was to determine which factors most significantly affect life expectancy in developing nations and to derive policy recommendations based on our findings.

@@

This finding also makes sense from a theoretical standpoint. The more educated an individual is, they will probably have more skills in adulthood. Therefore, they most likely will get better jobs compared to people with less education. As a result, those individuals will have more money for basic needs and healthcare increasing their overall life expectancy. Additionally, people who are more educated may also be more likely to be aware of health risks, which also increases their life expectancy. Hence, we recommend countries to prioritize increasing their education system compared to most of the other variables in the model.

@@

Reducing infant mortality, controlling HIV/AIDS and increasing healthcare expenditure are interconnected strategies that collectively raise life expectancy for developing countries. Our findings show that public health initiatives are crucial for improving infant healthcare, access to vaccinations, food banks and diseases prevention programs. Countries which aim to improve life expectancy should adopt a more comprehensive approach that combines investment in health care infrastructure which will have a trickle effect on all these other variables. Together, these efforts would counteract the drivers of infant mortality, HIV/AIDs and overall low life expectancy.

@@

@@

# Bibliography

>>>>

```
The first transformation we had was with health expenditure as a percentage of GDP per capita. For health expenditure, the figure above reveals a non-linear, positive exponential relationship with life expectancy. To linearize this relationship, we applied a $\log(1+Health Expenditure)$ transformation as the $\log$ function is in the inverse of exponential. We used $\log(1+x)$ rather than $\log(x)$ to accommodate the theoretical possibility of zero expenditure values since $\log(0$) is undefined.

@@

```
For infant deaths, we applied a $\log(1+\text{infant deaths})$ transformation based on the weak exponential relationship observed in the figure. We used $\log(1+x)$ to maintain a valid domain for potential zero values. The transformed data reveals a generally negative relationship with life expectancy, though with two distinct patterns: higher variability and occasional uniformity between 0-2, and a curved downward slope between 2-6. After testing multiple transformations, $\log(1+x)$ yielded the best results, despite not resulting in a perfectly linear relationship.

We also considered removing outlier countries with over 900 infant deaths per 1000 people as shown in the figure. This was done to potentially create a stronger exponential relationship between infant deaths and life expectancy. However, after removing the outliers, the transformation still showed no improved linearity after removal. Furthermore, the final model results remained largely unchanged without these outliers, indicating they were not influential. Therefore, we kept these data points in the final model dataset.

@@

```
For the third and final transformation, we decided to do a $\log(1+\text{HIV/AIDS Deaths})$ transformation. From the figure above, the relationship between aids and life expectancy is negative and exponential so a log transformation is valid. Similarly, $\log(1 + x)$ is used to ensure a valid domain. From the transformation plot above, $\log(1+\text{HIV/AIDS Deaths})$ and life expectancy appear to have a moderately strong negative linear relationship.

@@

In this study we explored factors that influence life expectancy in developing countries using multiple linear regression. We obtained our dataset from Kaggle [Life Expectancy WHO dataset](https://www.kaggle.com/datasets/kumarajarshi/life-expectancy-who), which compiles health, demographic, and economic indicators from the World Health Organization and United Nations. After cleaning the dataset and removing missing values, we focused on five predictors: schooling, alcohol consumption (categorical), infant deaths, HIV/AIDS prevalence, and health expenditure (as a percentage of GDP per capita). Because the relationships for infant deaths, HIV/AIDS, and health expenditure with life expectancy were all exponential-like, log-transformations were applied to stabilize variance and improve model fit. Residual analysis confirmed no major violations of linear regression model assumptions, while model selection using regsubsets confirmed that our final set of predictors balanced model simplicity and predictive strength. Our goal was to determine which factors most significantly affect life expectancy in developing nations and to derive policy recommendations based on our findings.

@@

@@

@@

## Recomendations For Developing Countries

Our findings related to reducing infant mortality, controlling HIV/AIDS and increasing healthcare expenditure are interconnected strategies that collectively raise life expectancy for developing countries. Therefore, public health initiatives are crucial for improving infant healthcare, access to vaccinations, food banks and diseases prevention programs. As a result, we recommend that countries which aim to improve life expectancy should adopt a more comprehensive approach that combines investment in health care infrastructure which will have a trickle effect on all these other variables. Together, these efforts would counteract the drivers of infant mortality, HIV/AIDs and overall low life expectancy.

Moreover, we have also determined that school is probably the factor that has the most significant impact on life expectancy, disregarding the counter-intuitive results of the alcohol consumption variable. Hence, we recommend countries prioritize increasing their education system compared to most of the other variables in the model. We also believe that increasing education may have a trickle-down effect on increasing life expectancy, from a theoretical standpoint. For instance, the more educated an individual is, the more likely they are to have more useful skills in adulthood. Therefore, they most likely will get better jobs compared to people with less education. As a result, those individuals will have more money for basic needs and healthcare, increasing their overall life expectancy. Additionally, people who are more educated may also be more likely to be aware of health risks, which also increases their life expectancy. Overall, increasing education in developing countries could theoretically have a large impact on increasing life expectancy in developing countries.

@@

# References
